Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-07-08 analysis/paper/paper.qmd
Updates of Discussion

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.([@willoughby1985; @Cabanes2024]).

Early descriptions of spheroids were provided by [@clark1955] and [@kleindienst1962], and [@leakey1971] later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan. Leakey’s definition of spheroids remains a widely cited typological standard: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex. Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests ([@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017]).

Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges. Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another. Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* : differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]). Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]). Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; \[\@mora2005\]; [@clark1955]). In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]). \[BM: yes, this is great, good job, this is a very important paragraph for the introduction of the paper. We might need to expand it a bit\]

@@

QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau. The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]). In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin. \[BM: great detail about the context, can we make a figure that includes a map and a photo of the site?\]

@@

A total of 955 stone artifacts were recovered from four localities at the QSY site. Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). \[BM: is it possible to use inline R code for these numbers?\] Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.

Raw materials at QSY were classified into four categories following [@pei2002] and [@pei2017]: lava, siliceous dolomite, chert, and others. Lava rock dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively). It seems that spheroids show no strong raw material preference. Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert. However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores. (pie chart in need?)

@@

Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the QSY facted spheroids includes the use of conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology ([@inizan1999], [@duke2021]).

The origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape ([@beyene2013]). Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control ([@wynn1995], [@stout2015]). Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production (exhausted cores), or as targeted end products of shaping strategies. In the case of QSY facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy ([@Muller2023], [@tittonSubspheroidsLithicAssemblage2020]). These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by [@mora2005], these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework ([@leakey1971]).

@@

>>>>

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].

Early descriptions of spheroids were provided by [@clark1955] and [@kleindienst1962], and [@leakey1971] later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan. Leakey’s definition of spheroids remains a widely cited typological standard: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex. Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].

Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges. Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another. Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* : differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979]. Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production [@sahnouni1997]. Another perspective holds that spheroids were naturally produced through high-intensity battering [@schick1994]; \[\@mora2005\]; [@clark1955]. In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy [@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]. \[BM: yes, this is great, good job, this is a very important paragraph for the introduction of the paper. We might need to expand it a bit\]

@@

```{r, echo=FALSE}
#| label: fig-QSY-geography-chronology
#| fig-cap: "Geological Context and Chrono-Stratigraphy of the QSY Site)"

library(here)
knitr::include_graphics(here("analysis/paper/figures/QSY_geography_chronology.png"))

QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau. The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [@li2000; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]. In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin [@yuwei2023; @ye2024], within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin (@fig-QSY-geography-chronology).

@@

A total of 955 stone artifacts were recovered from four localities at the QSY site. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; \@pei2017], the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). \[BM: is it possible to use inline R code for these numbers?\] Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.

Raw materials at QSY were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others. Lava rock dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively). It seems that spheroids show no strong raw material preference. Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert. However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores. (pie chart in need?)

@@

Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the QSY facted spheroids includes the use of conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].

The origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013]. Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015]. Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production (exhausted cores), or as targeted end products of shaping strategies. In the case of QSY facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by [@mora2005], these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework [@leakey1971].

@@

```{r tbl-3, echo=FALSE}

file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
data3 <- read_excel(file_path)

knitr::kable(data3, caption = "Summary of core typology and morphological parameters.")

site data: 1. Mujiaqiao, [@Wei1984]; 2. Liaohe river region, [@Li1991]; 3. Deshan Second Brick factory, [@Xi1994]; 4. Zhangjiatan, [@LixianMuseum1992]; 5. Huzhua Mount., [@Tan1999]; 6. Fangniu Mount., [@Fang2002]; 7. Fanba, [@Xia2022Fanba]; 8. Xiaokongshan, [@XiaokongshanTeam1988]; 9. Hejialiang, [@Wang2014Hanzhong]; 10. Longgangsi, [@Lu2006Liangshan]; 11. Shangdan basin, [@Wang2013Danjiang]; 12. Yaoshi basin, [@Wang2000Yaoshi]; 13. Choushuihe, [@Dai1973Lantian]; 14. Kehe, [@Jia1962Kehe]; 15. Shuigou, [@Huang1964Sanmenxia]; 16. Hezhigou, [@Liu1984Jingchuan]; 17. Xigou, [@Jia1959Xigou]; 18. Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19. Liujiacha, [@Xie1982Liujiacha]; 20. Dangcheng, [@Wu1989Dangcheng] ; 21. Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22. Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23. Jinsitai, [@Wang2010Jinstai]

```{r, echo=FALSE}
#| label: fig-Spheroids_distribution_China
#| fig-cap: "Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries"

library(here)
knitr::include_graphics(here("analysis/paper/figures/Spheroids_distribution_China.png"))

In China, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids_distribution_China. In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene. In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site in Inner Mongolia [@Wang2010Jinstai]. The widespread occurrence of spheroids across time and space may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.

over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China. However, that many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. By contrast, three sites—Xujiayao, Dingcun, and Jinsitai—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; \@SXA2014Dingcun]. Notably, the Xujiayao site alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

The raw materials used for spheroids at these archaeological sites primarily include quartz, quartzite, sandstone, and limestone, with occasional occurrences of lava and dolomite at a few localities. Among these, quartz and quartzite are the most widely used materials, found at 18 sites across China. In contrast, spheroids made of sandstone and limestone are mainly concentrated at sites distributed along the Fen River and Li River basins in Central China. Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun]. For example, at the Dingcun site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz. In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].

Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., Mujiaqiao, [@Wei1984]); pecking, where two blanks are struck together to round off edges—sometimes by splitting a cobble and using the halves against each other (e.g., Dingcun, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., Zhoukoudian, [@NIGPAS1985PekingMan]). Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids]. Many sites, such as Xujiayao and Dingcun, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan]. No clear relationship has yet been established between raw material types and specific production methods.

At Dingcun and Mujiaqiao, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun]. Mujiaqiao specimens retain parts of the cortex, while at Dingcun, flakes with large exterior and small interior platform angles were removed from spheroid blank to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984]. Of the six sites exhibiting similar strategies to QSY spehroids, all but Mujiaqiao are located within the Fen River and Lishui River regions. These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe]. The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy. This is a significant observation, as it implies that spheroid production techniques similar to those at QSY are more frequently associated with lithic assemblages characterized by intentional shaping strategies.

From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites. However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids_distribution_China). Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as Xujiayao, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology. Rather, we argue that faceted spheroids reflect a complex shaping strategy. Their presence within Acheulean or Developed Oldowan contexts is understandable, as these traditions share comparable levels of technological capability [@leakey1971; @Cabanes2024]. When faceted spheroids appear in more rudimentary industries or assemblages, it suggests that conceptual templates and advanced technological capacities were already present within those contexts [@tittonSubspheroidsLithicAssemblage2020].
